---
title: 'Data Challenge 4: Red Cards and Skin Tone'
output:
  html_notebook: default
  html_document: default
  pdf_document: default
---

#Introduction 

In this data challenge, we seek to answer the question “Are soccer referees more likely to give red cards to dark skin toned players than light skin toned players?”

The data set consists of information about players, information about the penalty cards they received, and information about the referees who issued the cards.

We also have information about the "implicit bias"" countries that the referees were from,

# Results
1. Some refs are biased
2. mean implicit bias is predictive of red cards
3. overall because many ref are used, no significant correlation with skin tone

```{r}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(knitr)
library(pscl)

```

```{r}
rc <- read.csv('redcard.csv')
head(rc)
```

#The Data

All of these data come from the same season, 2012-2013 so I think we can just look at the birth year for age. 
As a first pass, I will average the ratings for the two raters since they largely agree on skin tone. 

```{r rater variance}
qplot(rater1,rater2,data=rc)+
  geom_jitter()+
  ggtitle('How well did the two raters agree on skin tone?')+
  theme_fivethirtyeight()
```


```{r}

rc %>% 
  mutate(birthday = as.character(birthday),
         birthyear = substr(birthday,7,10),
         mean_rating = (rater1+rater2)/2) ->
  rcm
```

```{r}
head(rcm)
```

Variable Name | Description
------------- | -------------
meanIAT | mean implicit bias score (using the race IAT) for referee country
nIAT | sample size for race IAT in that particular country
seIAT | standard error for mean estimate of race IAT  
meanExp | mean explicit bias score (using a racial thermometer task) for referee country
nExp | sample size for explicit bias in that particular country
seExp |standard error for mean estimate of explicit bias measure

```{r}
rcm %>%
  group_by(playerShort) %>%
  summarize(position=unique(position))%>%
  ungroup() %>%
  summarize(missing_positions = sum(position == ''),
            total_players = n(),
            percent_missing = missing_positions/total_players) ->
  player_positions

kable(player_positions)
```
We are missing a lot of information about the positions of the players. 
We are also missing a lot of information about the skin tone of many players. 

# Results

There are two ways to look at this problem:
1. From the perspective of the players
2. from the persepctive of the referees

The question is posed in reference to the referees, but it will be useful to look at the players first.

##The Players

First let's make a plot to answer the question of whether there is a correlation between a players perceived skin tone and the number of red cards they receive:

```{r}


rcm %>%
  group_by(playerShort) %>%
  summarize(total_red_cards = sum(redCards),
            skin_tone = first(mean_rating)) %>%
  mutate(dark = skin_tone <= 0.25,
         light = skin_tone >= 0.75)->
  player_summary

player_summary$color = "mixed"
player_summary$color[is.na(player_summary$skin_tone)] = 'missing'
player_summary$color[player_summary$dark] = "dark"
player_summary$color[player_summary$light] = "light"
player_summary$color <- as.factor(player_summary$color)

qplot(skin_tone,data= player_summary)

qplot(total_red_cards, data= player_summary)

player_summary%>%
  filter(total_red_cards > 0) %>%
  ggplot(aes(skin_tone,total_red_cards))+
  geom_jitter()

player_summary%>%
  filter(total_red_cards > 6) %>%
  ggplot(aes(skin_tone,total_red_cards))+
  geom_jitter()+
  ggtitle('Players with more than 6 red cards')

player_summary %>%
  ggplot(aes(color,total_red_cards))+
  geom_boxplot()

player_summary %>%
  group_by(color) %>%
  tally()
```

We are missing skin tone information from 468/2053 (23%) of players. 
Of those with data, most were considered to have light skin by the raters (skin tone <= 0.25)
Most players did not receive any red cards during the season. Only a small number of players received more than 5 red cards in the season. 
All players who received more than 6 red cards were considered lighter than average. 


## Referees 

Now let's group by referees and see whether individual referees are more likely to give red cards to darker skinned players. 

```{r}
rcm %>%
  filter(!is.na(mean_rating)) %>%
  group_by(refNum) %>%
  summarize(total_reds_given = sum(redCards),
            total_reds_light = sum(redCards[mean_rating<= 0.25]),
            total_reds_dark = sum(redCards[mean_rating>= 0.75]))->
  ref_summary

qplot(total_reds_given,data=ref_summary)
    
ggplot(ref_summary %>% filter(total_reds_given > 0),
       aes(total_reds_dark, total_reds_light))+
  geom_hex()+
  ggtitle("Refs who gave at least one red card")
```
Most refs never give out any red cards.
Of those who do give out red cards, most give one red card to a player who was rated to have lighter skin and none to any players rated to have darker skin. 

However, there are 1123 players who were rated light and only 236 who were rated light. Can we adjust for this?

```{r}
ggplot(ref_summary %>% filter(total_reds_given > 0),
       aes(total_reds_dark, total_reds_light*4.7))+
  geom_hex()+
  geom_abline()+
  ggtitle("Refs who gave at least one red card")
```


```{r simple_regression}
rcm %>%
  mutate(binary_redCards = redCards >= 1) ->
  rcm
m <- glm(binary_redCards~mean_rating,data = rcm, family='binomial')
summary(m)
```

```{r}
player_summary %>% 
  mutate(any_red = total_red_cards >0) ->
  player_summary

m <- glm(any_red ~ color, data= player_summary, family = 'binomial')
summary(m)
```
```{r}

m <- glm(total_red_cards ~ dark, data= player_summary %>% 
           filter(color != 'missing'), family = 'poisson')
summary(m)
```
```{r zero-inflated}
#zero inflated poisson model?
m <- zeroinfl(total_red_cards ~ skin_tone, data= player_summary %>% 
           filter(color != 'missing'))
summary(m)
```
```{r}
total_red_cards
```


```{r}

m <- glm(any_red ~ skin_tone, data= player_summary %>% filter(color != 'missing'), family = 'binomial')
summary(m)
```

```{r}
rcm %>%
  filter(!is.na(mean_rating)) %>%
  group_by(refNum) %>%
  summarize(total_reds_given = sum(redCards),
            total_reds_light = sum(redCards[mean_rating<= 0.25]),
            total_reds_dark = sum(redCards[mean_rating>= 0.75]),
            meanIAT = first(meanIAT),
            meanExp = first(meanExp),
            dark_reds_proportion = total_reds_dark/total_reds_given) ->
  ref_summary

ref_summary$refNum<- as.factor(ref_summary$refNum)
```

```{r}
m <- glm(dark_reds_proportion ~ meanIAT,data = ref_summary, family = 'gaussian')
summary(m)
```

```{r}
m <- glm(dark_reds_proportion ~ meanExp,data = ref_summary, family = 'gaussian')
summary(m)
```

```{r}

```

